%% Adapted from GaTech CS 7545 Machine Learning Theory given by Prof. Jacob Abernathy Fall 2018. Original latex file: https://github.com/mltheory/CS7545/blob/master/scribe/CS7545scribe_template.tex



%%%%%PLEASE CONSIDER CORRECTIONS AT PLACES INDICATED%%%%%%%%
\documentclass{article}
%%%%%Packages Used, add more if necessary%%%%
\usepackage{amsmath,amsfonts,amssymb,graphicx,fullpage, float}
\setlength{\topmargin}{-0.6 in}
\setlength{\textheight}{8.5 in}
\setlength{\headsep}{0.75 in}
\usepackage{mdframed}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{hyperref}

%%%%% NO NEED TO EDIT THIS PREAMBLE %%%%%%%%
%%% PREAMBLE %%%
\newcounter{lecnum}
\renewcommand{\thepage}{\thelecnum-\arabic{page}}
\renewcommand{\thesection}{\thelecnum.\arabic{section}}
\renewcommand{\theequation}{\thelecnum.\arabic{equation}}
\renewcommand{\thefigure}{\thelecnum.\arabic{figure}}
\renewcommand{\thetable}{\thelecnum.\arabic{table}}
\newcommand{\lecture}[4]{
   \pagestyle{myheadings}
   \thispagestyle{plain}
   \newpage
   \setcounter{lecnum}{#1}
   \setcounter{page}{1}
   \noindent
   \begin{center}
   \framebox{
      \vbox{\vspace{2mm}

      \hbox to 6.28in {
      {\bf ECE 4252/6252 (FunML) Fundamentals of Machine Learning \hfill 2021--2028} }

      \vspace{2mm}
      \hbox to 6.28in {
      {Courses developed by Professor Ghassan AlRegib
      (\href{https://alregib.ece.gatech.edu/}{alregib.ece.gatech.edu})
      \hfill For inquiries: \href{mailto:alregib@gatech.edu}{alregib@gatech.edu}} }

      \vspace{3mm}
      \hbox to 6.28in { {\Large \hfill Lecture #1: #2 \hfill} }

      \vspace{2mm}
      \hbox to 6.28in { {\it Co-Instructors: #3 \hfill #4} }

      \vspace{2mm}}
   }
   \end{center}

   % -------- PEOPLE OUTSIDE BOX --------
   \vspace{2mm}
   \noindent{\bf Contributors:} Dr. Ahmad Mustafa, Dr. Motaz Alfarraj, Dr. Ashraf Alattar, Dr. Chen Zhou

   \vspace{1mm}
   \noindent{\bf Teaching Assistants} with remarkable contributions include: Kuo-Wei Lai, Wuyang Du, Shiva Mahato, Michael Zhou, Ninghan Zhong

   \markboth{Lecture #1: #2}{Lecture #1: #2}

   \vspace{2mm}
\noindent {\bf Disclaimer}:
{All content of these notes are part of this course at Georgia Tech. Any re-use or distribution is not permitted without pre-approved permission. All these notes belong to, created by, and copyrighted for Ghassan AlRegib and Mohit Prabhushankar, Georgia Tech, 2021--2028.}

\vspace{2mm}
\noindent {\bf License}:
{These lecture notes are licensed under the
\href{https://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License}.}

   \vspace{2mm}
   \noindent {\bf Errata}:
   {\it Please submit any errata you find using the following form:
   \href{https://forms.office.com/r/fbg9dMWPgY}{Errata Form for FunML Textbook}
   or visit: \url{https://forms.office.com/r/fbg9dMWPgY}}

   \vspace*{4mm}
}
\renewcommand{\cite}[1]{[#1]}
\def\beginrefs{\begin{list}
        {[\arabic{equation}]}{\usecounter{equation}
         \setlength{\leftmargin}{2.0truecm}\setlength{\labelsep}{0.4truecm}%
         \setlength{\labelwidth}{1.6truecm}}}
\def\endrefs{\end{list}}
\def\bibentry#1{\item[\hbox{[#1]}]}

\newcommand{\challenge}[2]{\noindent \textbf{(Challenge Problem)} \emph{#1}: #2 }

\newcommand{\exercise}[1]{\noindent \textbf{(Exercise)} #1 }

\newcommand{\fig}[3]{
			\vspace{#2}
			\begin{center}
			Figure \thelecnum.#1:~#3
			\end{center}
	}
%%% END_OF_PREAMBLE %%%



%%%%You may add more \newtheorem if necessary%%%%%%
\newtheorem{theorem}{Theorem}[lecnum]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newenvironment{proof}{{\bf Proof:}}{\hfill\rule{2mm}{2mm}}


\newcommand{\dom}{\mathrm{dom}}
\begin{document}

\lecture{29}{Vision-Language Models}{Ghassan AlRegib and Mohit Prabhushankar}{}%

\section{Lecture Objectives}

This lecture extends the study of Large Language Models to the multimodal setting where visual and linguistic signals are jointly processed. We begin by formalizing the vision-language alignment problem and motivating it with downstream tasks such as image captioning, visual question answering, and zero-shot image classification. We study contrastive pretraining with CLIP as the foundation for aligning image and text representation spaces at scale, and analyze how contrastive objectives produce transferable embeddings. We then examine architectural designs that bridge a pretrained visual encoder with a pretrained language model, including the cross-attention gating approach of Flamingo, the projection-based connector of LLaVA, and the Q-Former bridge of BLIP-2. We study the Visual Transformer (ViT) in detail as the canonical image encoder, discussing how images are patchified and fed into a standard Transformer. We discuss multimodal instruction tuning, which extends LLM-style SFT and RLHF to image-conditioned settings, and survey benchmark tasks used to evaluate VLMs. We close with an analysis of failure modes unique to multimodal models, including hallucination of visual content, modality imbalance, and challenges in spatial and relational reasoning \cite{1,2,3,4}.

\section{Motivation and Problem Setting}

Human perception is inherently multimodal: we seamlessly integrate visual, auditory, and linguistic information when understanding the world. A Vision-Language Model (VLM) attempts to replicate this cross-modal integration in a machine learning system. Formally, a VLM takes as input an image $I \in \mathbb{R}^{H \times W \times C}$ and a text query $q = (w_1, \ldots, w_T)$, and produces a text response:
\begin{equation}
P(y \mid I, q) = \prod_{t} P(y_t \mid I, q, y_1, \ldots, y_{t-1}).
\end{equation}

Downstream applications include:
\begin{itemize}
  \item \textbf{Image captioning:} Generate a natural-language description of an image.
  \item \textbf{Visual Question Answering (VQA):} Answer a free-form question about an image.
  \item \textbf{Zero-shot image classification:} Rank candidate class names by their compatibility with an image embedding.
  \item \textbf{Optical Character Recognition (OCR) and document understanding:} Parse and reason over text embedded in images.
  \item \textbf{Grounding:} Localize objects or regions in an image referred to by a textual expression.
  \item \textbf{Multimodal instruction following:} Perform multi-step visual tasks described in natural language.
\end{itemize}

\section{The Vision Transformer (ViT)}

\subsection{Patchification and Embedding}

A Vision Transformer (ViT) \cite{5} processes an image by dividing it into a grid of non-overlapping patches of size $P \times P$ pixels. For an image of resolution $H \times W$, this yields $N = (H \cdot W) / P^2$ patches. Each patch $\mathbf{p}_i \in \mathbb{R}^{P \times P \times C}$ is flattened and linearly projected into a $d$-dimensional embedding:
\begin{equation}
\mathbf{z}_i = \mathbf{W}_\text{patch}\, \text{vec}(\mathbf{p}_i) + \mathbf{b},
\end{equation}
where $\mathbf{W}_\text{patch} \in \mathbb{R}^{d \times P^2 C}$. A learnable class token $[\text{CLS}]$ is prepended, and 1D positional embeddings are added to each patch embedding. The resulting sequence of $N+1$ tokens is processed by a standard Transformer encoder.

\subsection{Self-Attention over Image Patches}

Because ViT uses the same Transformer attention mechanism as text models, each patch can attend to every other patch in the image. This global receptive field distinguishes ViT from convolutional networks, which have a local receptive field that grows with depth. The $[\text{CLS}]$ token aggregates information from all patches through attention and serves as the global image representation used for downstream tasks.

ViT requires large datasets to achieve strong performance; its lack of inductive biases (spatial locality, translation equivariance) compared to CNNs is a disadvantage on small datasets but an advantage when trained on hundreds of millions of image-text pairs, where the model can learn these biases from data.

\subsection{Patch Resolution and Scale}

Common ViT configurations include:
\begin{itemize}
  \item \textbf{ViT-B/16:} Base model, $16 \times 16$ patches, 86M parameters.
  \item \textbf{ViT-L/14:} Large model, $14 \times 14$ patches, 307M parameters.
  \item \textbf{ViT-G/14:} Giant model, $14 \times 14$ patches, 1.8B parameters (used in CLIP and Flamingo).
\end{itemize}
Smaller patch size increases the number of tokens (quadratically increasing attention cost) but preserves finer spatial detail.

\section{Contrastive Language-Image Pretraining (CLIP)}

\subsection{The Contrastive Objective}

CLIP \cite{1} (Contrastive Language-Image Pre-Training) learns aligned image and text representations by training on 400 million image-text pairs collected from the internet. The model consists of two encoders:
\begin{itemize}
  \item An \textbf{image encoder} $f_\theta: \mathbb{R}^{H \times W \times 3} \to \mathbb{R}^d$ (a ViT or ResNet).
  \item A \textbf{text encoder} $g_\phi: \mathcal{V}^* \to \mathbb{R}^d$ (a Transformer).
\end{itemize}
For a batch of $B$ image-text pairs $\{(I_i, t_i)\}_{i=1}^B$, image embeddings $\mathbf{v}_i = f_\theta(I_i) / \|f_\theta(I_i)\|$ and text embeddings $\mathbf{u}_i = g_\phi(t_i) / \|g_\phi(t_i)\|$ are computed. The similarity matrix $\mathbf{S} \in \mathbb{R}^{B \times B}$ has entries:
\begin{equation}
S_{ij} = \tau \cdot \mathbf{v}_i^\top \mathbf{u}_j,
\end{equation}
where $\tau$ is a learned temperature parameter. The training loss is the sum of two symmetric cross-entropy terms:
\begin{equation}
\mathcal{L}_\text{CLIP} = -\frac{1}{2B}\sum_{i=1}^B \left[\log \frac{e^{S_{ii}}}{\sum_j e^{S_{ij}}} + \log \frac{e^{S_{ii}}}{\sum_j e^{S_{ji}}}\right].
\end{equation}
Each term treats the matching pair $(I_i, t_i)$ as a positive and all other $B-1$ pairs as negatives. Training pushes matched image-text embeddings together and unmatched embeddings apart in the shared $d$-dimensional space.

\subsection{Zero-Shot Classification with CLIP}

After training, CLIP can classify images into arbitrary categories without any task-specific training. Given a set of class names $\{c_1, c_2, \ldots, c_K\}$, text prompts are constructed (e.g., ``a photo of a \{class\}'') and encoded with the text encoder. An image is then classified to the class with highest cosine similarity:
\begin{equation}
\hat{y} = \arg\max_{k} \mathbf{v}^\top \mathbf{u}_{c_k}.
\end{equation}
On ImageNet, CLIP achieves accuracy comparable to supervised ResNet-50 without any labeled training examples. This demonstrates that the joint visual-linguistic embedding space generalizes well to novel classification tasks.

\subsection{CLIP as a Visual Backbone}

Because CLIP visual encoders are trained on diverse image-text pairs, they produce rich, transferable image representations. Modern VLMs almost universally use a pretrained CLIP ViT as the visual backbone, fine-tuning it lightly or keeping it frozen while training the language model interface.

\section{Architectures for Vision-Language Models}

A core design decision in building a VLM is how to connect the visual encoder to the language model. Three main architectural patterns have emerged.

\subsection{Cross-Attention Gating: Flamingo}

Flamingo \cite{2} inserts \emph{gated cross-attention} layers between every layer of a frozen, pretrained language model. Each cross-attention layer allows the language model's hidden states $\mathbf{H}$ to attend over visual features $\mathbf{V}$ extracted by a frozen ViT:
\begin{equation}
\mathbf{H}' = \mathbf{H} + \tanh(\alpha) \cdot \text{CrossAttention}(\mathbf{H}, \mathbf{V}, \mathbf{V}),
\end{equation}
where $\alpha$ is a learnable scalar initialized to 0, ensuring that training begins with the language model unchanged. A Perceiver Resampler module compresses the potentially large set of visual tokens into a fixed number of learned query tokens, making the approach efficient for high-resolution images and videos.

Flamingo demonstrates strong few-shot performance on VQA and captioning benchmarks when provided with visual-linguistic examples in context, analogous to LLM few-shot learning.

\subsection{Projection Connector: LLaVA}

LLaVA (Large Language and Vision Assistant) \cite{3} takes a simpler approach: a \emph{linear projection} (or small MLP) maps visual features from the CLIP encoder directly into the token embedding space of a pretrained language model (e.g., LLaMA or Vicuna):
\begin{equation}
\mathbf{H}_v = \mathbf{W}\, f_\theta(I),
\end{equation}
where $\mathbf{H}_v \in \mathbb{R}^{N \times d_\text{LLM}}$ are the projected visual tokens. These are concatenated with the text token embeddings and fed as a single sequence into the language model. The LLM is then instruction-tuned on visual instruction-following data using standard SFT.

LLaVA demonstrates that a simple projection layer combined with high-quality multimodal instruction data can produce competitive VLMs with much lower training cost than architecturally complex alternatives.

\subsection{Q-Former Bridge: BLIP-2}

BLIP-2 \cite{4} introduces a \emph{Query Transformer (Q-Former)}, a lightweight Transformer with $K$ learnable query tokens that extract information from the frozen visual encoder via cross-attention:
\begin{equation}
\mathbf{Q}' = \text{QFormer}(\mathbf{Q}, f_\theta(I)),
\end{equation}
where $\mathbf{Q} \in \mathbb{R}^{K \times d}$ are the learnable queries and $f_\theta(I)$ are the frozen image features. The $K$ output vectors (typically $K = 32$, much smaller than the $N$ patch tokens) are projected into the LLM embedding space and prepended to the text tokens.

The Q-Former acts as an \emph{information bottleneck}: it forces the $K$ queries to distill the most task-relevant visual information from the image features. This compression reduces the number of visual tokens fed into the expensive LLM, lowering computational cost while retaining key visual details.

\subsection{Comparison of Architectural Patterns}

\begin{center}
\begin{tabular}{lccc}
\hline
\textbf{Approach} & \textbf{Visual tokens to LLM} & \textbf{LLM frozen?} & \textbf{Training cost} \\
\hline
Flamingo (cross-attn) & Resampled ($\sim$64) & Yes & High \\
LLaVA (projection) & All patch tokens ($N$) & No (SFT) & Medium \\
BLIP-2 (Q-Former) & Fixed $K$ ($\sim$32) & Yes & Low \\
\hline
\end{tabular}
\end{center}

\section{Multimodal Instruction Tuning}

Pre-trained VLMs that are only trained for captioning or image-text matching do not follow free-form instructions well. Multimodal instruction tuning fine-tunes the VLM on datasets of the form:
\[
(\text{image}, \text{instruction}, \text{response}),
\]
using the same SFT objective as for LLMs. Data is typically generated by prompting a capable LLM (e.g., GPT-4) with image captions and region descriptions to produce diverse question-answer pairs that require visual reasoning. Examples include:
\begin{itemize}
  \item ``Describe the relationship between the two people in the image.''
  \item ``What is unusual about this scene?''
  \item ``How many objects of the same color as the chair are in the image?''
\end{itemize}
LLaVA-Instruct-158K is a widely-used dataset of 158K such examples. Models fine-tuned on multimodal instruction data can engage in multi-turn visual conversations and perform complex reasoning over images.

\subsection{RLHF for VLMs}

Reinforcement Learning from Human Feedback can be applied to VLMs in the same way as for LLMs, with the reward model conditioned on both the image and the text response. This addresses failure modes specific to vision-language settings, such as describing objects that are not present in the image (visual hallucination). Multimodal RLHF is an active research direction; recent work shows that preference feedback on visual outputs reduces hallucination rates substantially.

\section{Evaluation Benchmarks}

\subsection{Visual Question Answering}

\textbf{VQA v2} is a standard benchmark with 265K question-answer pairs over 200K images. Questions require recognizing objects, counting, reading text, and spatial reasoning. Performance is measured as accuracy over a pool of 10 human-collected answers per question. State-of-the-art VLMs achieve 70--85\% on VQA v2.

\textbf{GQA} (Graph Question Answering) provides compositional visual questions over structured scene graphs, testing whether models can perform multi-step reasoning over object attributes and relations.

\subsection{Image Captioning}

\textbf{COCO Captions} is evaluated with BLEU, METEOR, ROUGE, CIDEr, and SPICE metrics that measure the n-gram overlap and semantic similarity between generated and reference captions. BLEU-4 and CIDEr are most commonly reported. However, these metrics correlate imperfectly with human judgment; recent work uses LLM-based evaluators.

\subsection{Multimodal Benchmarks}

\begin{itemize}
  \item \textbf{MMBench / MME:} Comprehensive evaluations of VLM capabilities across perception, cognition, reasoning, and hallucination.
  \item \textbf{ScienceQA:} Multi-choice science questions with diagrams, testing scientific reasoning.
  \item \textbf{TextVQA / DocVQA:} Questions about text appearing in natural images and scanned documents.
  \item \textbf{POPE:} Probing Object Hallucination Evaluation --- determines whether a VLM hallucinates objects not present in the image.
\end{itemize}

\section{Failure Modes and Limitations}

\subsection{Visual Hallucination}

The most salient failure mode of VLMs is \emph{visual hallucination}: generating text that is fluent and contextually plausible but describes visual content not present in the image. This occurs because the language model component has strong textual priors from LLM pretraining that can override visual evidence. For example, a model trained on many images of kitchens may describe a microwave in an image even when none is present.

Mitigation strategies include: (1) object-level grounding supervision, which requires the model to justify descriptions with detected regions; (2) RLHF with hallucination-sensitive reward signals; and (3) chain-of-thought prompting that forces explicit reference to image evidence.

\subsection{Modality Imbalance}

During multimodal fine-tuning, the language model component often dominates because its weights are pre-trained on far more data than the visual component. The model may answer visual questions using only textual priors without truly processing the image. Diagnosing modality imbalance involves testing model performance when the image is replaced with a blank or random image.

\subsection{Spatial and Relational Reasoning}

VLMs trained primarily on captioning and VQA tend to perform poorly on tasks requiring precise spatial reasoning (``Is the red box to the left or right of the blue circle?''), counting, and compositional attribute binding (``What is the color of the largest object?''). These limitations reflect the fact that patch-based Transformer encoders aggregate spatial information through attention, losing explicit geometric structure.

\subsection{Context Length and Resolution}

High-resolution images produce many patch tokens, quickly filling the context window of the language model. Techniques such as dynamic resolution (splitting images into tiles), token merging, and Q-Former compression help manage context length at the cost of fine-grained detail.

\section{Summary}

Vision-Language Models combine a pretrained visual encoder (typically a ViT trained with CLIP-style contrastive objectives) with a pretrained language model through an architectural connector (projection, cross-attention gating, or Q-Former). CLIP provides a strong foundation by aligning image and text representations in a shared embedding space via large-scale contrastive pretraining. Architectural choices in connecting modalities involve a trade-off between expressivity, computational cost, and the degree of joint fine-tuning. Multimodal instruction tuning adapts the pretrained VLM to follow diverse visual instructions, and RLHF reduces hallucination. Key open challenges include visual hallucination, spatial reasoning, modality imbalance, and scaling to higher resolutions and longer contexts.

\section*{Questions and Answers}

\begin{mdframed}
\textbf{Q: Why is contrastive pretraining (CLIP) preferred over generative pretraining for the visual encoder?}

\textbf{A:} Contrastive pretraining aligns image representations with language representations directly --- the learned visual features are structured to be compatible with text descriptors, which is precisely the property needed for downstream VQA and captioning. Generative visual models (e.g., VAEs, GANs, MAE) optimize for reconstructing pixel values and tend to encode low-level visual statistics. CLIP-trained encoders encode semantically meaningful, language-aligned features that transfer better to vision-language tasks, especially in zero-shot and few-shot settings.
\end{mdframed}

\begin{mdframed}
\textbf{Q: What is the role of the Q-Former in BLIP-2, and why is it useful?}

\textbf{A:} The Q-Former is a Transformer with $K$ learnable query tokens that attend over image features via cross-attention. Its role is to compress the $N$ patch tokens from the visual encoder into a fixed, small number of $K$ output vectors that capture the most task-relevant visual information. This is useful for two reasons: (1) it reduces the number of visual tokens fed into the language model from $N$ (often 256--1024) to $K$ (typically 32), keeping the LLM's context window manageable; (2) the queries act as an adaptive summarization mechanism that can learn to extract different aspects of the image depending on the textual context.
\end{mdframed}

\begin{mdframed}
\textbf{Q: How does zero-shot image classification work with CLIP?}

\textbf{A:} CLIP encodes each candidate class name into a text embedding using a prompt template such as ``a photo of a [class name].'' The query image is encoded by the visual encoder. The predicted class is the one whose text embedding has the highest cosine similarity to the image embedding. No task-specific labeled data is needed. The zero-shot capability emerges because CLIP's training objective forces image and text embeddings to be aligned in the same metric space, so an image of a cat and the text ``a photo of a cat'' end up nearby, while ``a photo of a dog'' is farther away.
\end{mdframed}

\begin{mdframed}
\textbf{Q: What is visual hallucination and how does it differ from LLM hallucination?}

\textbf{A:} Both LLM and VLM hallucination refer to generating confident but factually incorrect content. In LLMs, hallucination typically concerns factual claims about the world (wrong dates, nonexistent citations). In VLMs, visual hallucination specifically refers to describing image content that is not present in the given image --- the model generates plausible visual descriptions based on language priors rather than actual visual evidence. Visual hallucination is particularly problematic because it is harder to detect (the generated text is syntactically correct and contextually plausible) and can occur even when the model has correctly encoded the image, if the language decoder overrides visual signals.
\end{mdframed}

\begin{thebibliography}{9}
\bibitem{1} Radford, A., et al. ``Learning Transferable Visual Models From Natural Language Supervision.'' \emph{ICML}, 2021.
\bibitem{2} Alayrac, J.-B., et al. ``Flamingo: a Visual Language Model for Few-Shot Learning.'' \emph{NeurIPS}, 2022.
\bibitem{3} Liu, H., et al. ``Visual Instruction Tuning.'' \emph{NeurIPS}, 2023.
\bibitem{4} Li, J., et al. ``BLIP-2: Bootstrapping Language-Image Pre-training with Frozen Image Encoders and Large Language Models.'' \emph{ICML}, 2023.
\bibitem{5} Dosovitskiy, A., et al. ``An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale.'' \emph{ICLR}, 2021.
\end{thebibliography}

\end{document}
